Migration of Gas Giant Planets in Gravitationally Unstable Disks

### 2011-06-23 Text/migration.tex
Final, Final, Final edits. Added Dick's comment and submitted 06/07/2011
	modified:   Text/migration.tex
	modified:   Text/migration_apjlsubmission.tex

git-svn-id: http://jaga.astro.indiana.edu/paperrepo/ApJL/Migration2010@87 3b7f7548-5c66-0410-83ab-3997167635fc
Changes prompted by first review by referee
Added paragraph in discussion section; Added h/r aspect ratio plot for all runs at 2 different times.

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state (EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. Emerging studies, such as \citet{baruteau2011} are just beginning to address some of these issues. 
\citet{boss2005} and \citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim 0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur  \citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU) \citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part on their radial migration, which is a chaotic and messy affair in a fragmenting disk \citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves, such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

The model disk, based on an equlibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d = 0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim \varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007} minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3 ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. This is a reasonable estimate for comparison purposes, because our resolution is probably insufficient to compute the corotation torque \citep{paardekooper2010}. 
While the interaction with a GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in direction than in the $t = 0$ planet run. The important departures from laminar disk theory are threefold: the more erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$ type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

@@

@@

%***DICK: Scott, remember that figures need some improvements.

\begin{figure}[t]
\center
\includegraphics[width=11cm]{Figures/amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$ as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet run, respectively.}
\label{fig:Am}
\end{figure}

>>>>

Migration of gas giant planets due to interactions with a circumstellar gas disk can play a major role in defining the
architecture of planetary systems. Work on migration \citep[see review by][]{papaloizou2007} has included gravitational
interaction of planets with both laminar and turbulent disks. However, radiative transport, detailed equations of state
(EOS), and the self-gravity of the gas disk have usually been ignored; the effects of a non-isothermal EOS have only
recently been included \citep[e.g.,][]{paardekooper2006,paardekooper2010,paardekooper2011}. Emerging studies, such as
\citet{baruteau2011} and the one reported here, are beginning to address some of these issues.  \citet{boss2005} and
\citet{mayer2004} examined radial migration of planet-mass fragments in gravitationally unstable disks, but their disks
were violently disrupted by fragmentation under conditions (radii $<$ 40 AU, disk-to-star mass ratios $M_d/M_s \sim
0.1$, and stellar mass $M_s \sim 1\;M_{\odot}$) where fragmentation may not actually occur
\citep{rafikov2005,rafikov2007,boley2006,boley2007b,boley2008,forgan2009,cai2010}. More recently, fragmentation into
clumps with gas giant or brown dwarf masses has been documented in numerical simulations of disks that are relatively
massive ($M_d/M_s \sim$ a few tenths) and spatially extended (outer radii $>$ 50 AU)
\citep{krumholz2007,stamatellos2007,stamatellos2009,boley2009,boley2010}, where fragmentation is expected from
semi-analytic arguments \citep[e.g.,][]{clarke2009,rafikov2009,dodson2009}. The fate of the clumps then depends in part
on their radial migration, which is a chaotic and messy affair in a fragmenting disk
\citep[e.g.,][]{boley2009,boley2010,vorobyov2010,boley2010b}. The occurrence of gravitational instabilities (GIs) may be
episodic \citep[e.g.,][]{vorobyov2006,vorobyov2010b,zhu2010}. Clumps that survive and contract to the dimensions of
young planets can later find themselves in a disk that erupts again into GI activity. As the star/disk system evolves,
such a protoplanet may end up in a region of a GI-active disk where fragmentation does not occur.

@@

The model disk, based on an equilibrium model from \citet{pickett2003}, orbits a 1 $M_{\odot}$ star and has a mass $M_d =
0.14 M_{\odot}$, inner and outer radii at 5 and 40 AU, and an initial surface density distribution $\Sigma \sim
\varpi^{-1/2}$. The time unit of one ORP (= outer rotation period) is defined as the rotation period of the initial disk
at $\varpi \approx$ 32 AU, or about 180 yr. The disk is initially located between radial grid zones 30 and 240 and is
close to isentropic, which results in a Toomre-$Q$ distribution with a marginally unstable \citep[see ][]{durisen2007}
minimum value of 1.38 at radial grid zone 161 (26.7 AU). The computational grid extends radially to 512 zones to
accommodate expansion of the outer disk once GIs become nonlinear. An outflow boundary condition is enforced at the
upper vertical grid boundary, the outer radial grid boundary, and an inner radial boundary at 2 AU. To seed
nonaxisymmetry, the density distribution is given an initial 0.01 \% random cell-to-cell perturbation.

@@

Figure \ref{fig:a} suggests that the overall radial drift during the first 10 ORP of the $t = 0$ planet run is not too
different from what one would get by extrapolating the migration due solely to the planet's wake during the first 2 or 3
ORP, where the average radial migration is about 4$\times10^{-3}$ AU yr$^{-1}$. Although there are no simple formulae
that are fully applicable to our radiatively cooled disk, using eq. (70) from \citet{tanaka2002} with parameters from
the simulated disk, we estimate 6$\times10^{-3}$ AU yr$^{-1}$ as an analytic expectation. While the interaction with a
GI-active disk does not seem to result in a drastically different overall migration rate, the direction and magnitude of
migration in both simulations is quite variable and seems to depend strongly on interactions of the planet with discrete
modes in the disk. In $t = 10$ planet run, the overall radial migration rate is similar but is even more erratic in
direction than in the $t = 0$ planet run. The important departures from laminar disk theory are threefold: the more
erratic nature of migration, the importance of interactions with discrete low-order modes, and stalling of migration
near the inner edge of the strongest GI activity. The migration is neither a random walk nor monotonic but has a chaotic
character due to nonlinear interactions. For all these reasons, it will be important in future studies to explore the
dependence of the planet's behavior on its mass and on its placement relative to the site of GI-eruptions (for $t = 0$
type runs) and relative to the CRs of low-order GI modes (for $t = 10$ type runs).

@@

Due to the global nature of the GIs and their strong interaction with the embedded planet, it is important that the gas
phase physics are as accurate as possible and detailed analysis of the interaction between the planet and global modes
is required. Of particular importance is the treatment of cooling in the disk. Although these simulations show some
similarities to \citet{baruteau2011}, their simulations employ an idealized cooling rate governed by the local dynamical
timescale while the simulations in this work use the more realistic radiative physics of \citet{boley2007b}. {\bf
  citations on global vs. local needed} Another primary difference between this work and that of \citet{baruteau2011} is
the simulations presented here are three dimensional. Many studies have shown that three dimensional simulations are
necessary to accurately measure the effects of GIs {\bf need some more citations here}. Further in depth analysis of the
GIs and the interaction between the GIs and the planet will be presented in a future work.

{\bf Say something about possible physical scenarios for t=0 run}

@@

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/hr_migration.eps}
\caption{(Color figure available in the online version) Plots of the disk aspect ratio $h/r$ as determined by a decrease
  from midplane density by a factor of $e^2$ for the fiducial and $t=0$ runs (black) at $\approx 10$ ORP and the fiducial and
  $t=10$ runs (red) at $\approx 20$ ORP.}
\label{fig:hr}
\end{figure}

\begin{figure}[t]
\center
\includegraphics[width=12cm]{Figures/amt.eps}
\caption{(Color figure available in the online version) Global amplitudes of nonaxisymmetric density perturbations $A_m$
  as a function of time for individual cos($m\phi$) perturbations. The formula for computing $A_m$ is equation (15) of
  \citep{boley2006}. Here the integrals are done only over the disk outside 15 AU to suppress contributions from edge
  effects in the inner disk. The top and bottom panels correspond to the fiducial run (no planet) and the $t = 0$ planet
  run, respectively.}
\label{fig:Am}
\end{figure}
